---
# title: "Lecture 01"
# author: "Kasthuri Kannan"
# date: "Sept 07, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

### Homework Guidelines

---

##### Github pre-requisites 

The following are pre-requisites for submitting the homework.

* We assume you already have a Github account and you are able to push/pull into Github. If not, please read [Lecture 01](https://kannan-kasthuri.github.io/Lecture_Files/Lecture_01.html).

* Claim your repository in Github - otherwise ask me and I will make you one.

---

##### README.txt

As your first homework, you will create a README.txt file and push it into your repository. You need to constantly update this README.txt file as you work through the rest of the homeworks.

---

##### Naming files

Throughout this course you will create R Markdown files and render it as html which you will push to your coursework repository. Your markdown files **should be named** with an extension ".Rmd" or ".rmd". RStudio’s “Knit to HTML” button will compile the document for you to actual HTML and open a preview. 
You may want to read the chapter [R Markdown](http://r4ds.had.co.nz/r-markdown.html) in the book R for Data Science if you want to get started before we cover this in the class.

Make sure you organize your files really well and name appropriately so it is easier for us to grade. Part of the grades will go for your ability to organize files for easier access and readability. 

---

##### How to submit homework?

1. Make sure you have saved the work in your local directory. When you hit the save icon in the RStudio editor, you will see the markdown .Rmd file popping up in the Git panel of RStudio, ready to be committed.

2. Click "Knit to HTML" in the RStudio editor to render the associated HTML file. A preview of the HTML file should pop up. At this point you will see two files, the markdown and the html documents, ready to be committed to Git. Check the files.

```{r, echo=FALSE}
htmltools::includeHTML("../Figures/Fig_2.html")
```
<br>

3. Commit the work to the Git repository.

4. Push your work into Github.

5. Indicate that you have completed your work by creating a new issue in our [discussion page](https://github.com/PDA-NYU-Fall-2017/Discussion/issues) with the "latest commit" information for your .Rmd and .html files and submit it. You are done now.

---
