---
# title: "Lecture 02"
# author: "Kasthuri Kannan"
output: html_document

---

<style>
body {
text-align: justify}
</style>

### An Introduction to Optimization (Oct. 23, 2017)

---

Optimization is a graduate level course for a semester or two, typically in engineering and mathematics departments. Here we will be giving an overview of the subject, in particular on how the concepts fit into other areas such as machine learning, image processing etc.

We will see the following:

1. What optimization is used for?

2. Different optimization concepts.

3. Commonly used terminologies.

4. Basic pointer on how to perform optimization with programming languages such as R.

based on an overview of the following concepts:

a. Likelihood and maximum likelihood estimates
b. Cost functions
c. Gradient descent and minimas
d. Convex/Non-convex and differentiable functions
e. Stochastic gradient descent
f. Regularization, sparse coding and momentum methods

---

#### Likelihood and maximum likelihood estimates

---

Likelihood is the probablity of observing the data given a model. In the univariate case, let us say that we have a set of data $X = \{x_1,x_2,\dots,x_n\}$ and model with parameter $\gamma$, then the likelihood is defined as,

$$ L(\gamma \mid X) = P(X \mid \gamma)$$

If our observations are independent and identically distributed, then we can treat this as product of probabilities of observing each individual data point. That is,

$$ L(\gamma \mid X) = \prod_{i=1}^{n} P(x_{i} \mid \gamma)$$

This may seem really abstract and so we can look at concrete example.

---

Let us work with the HANES data set.

```{r eval=FALSE, message=FALSE, echo = TRUE}
  # Load the package RCurl
  library(RCurl)
  # Import the HANES data set from GitHub; break the string into two for readability
  # (Please note this readability aspect very carefully)
  URL_text_1 <- "https://raw.githubusercontent.com/kannan-kasthuri/kannan-kasthuri.github.io"
  URL_text_2 <- "/master/Datasets/HANES/NYC_HANES_DIAB.csv"
  # Paste it to constitute a single URL 
  URL <- paste(URL_text_1,URL_text_2, sep="")
  HANES <- read.csv(text=getURL(URL))
  # Rename the GENDER factor for identification
  HANES$GENDER <- factor(HANES$GENDER, labels=c("M","F"))
  # Rename the AGEGROUP factor for identification
  HANES$AGEGROUP <- factor(HANES$AGEGROUP, labels=c("20-39","40-59","60+"))
  # Rename the HSQ_1 factor for identification
  HANES$HSQ_1 <- factor(HANES$HSQ_1, labels=c("Excellent","Very Good","Good", "Fair", "Poor"))
  # Rename the DX_DBTS as a factor
  HANES$DX_DBTS <- factor(HANES$DX_DBTS, labels=c("DIAB","DIAB NO_DX","NO DIAB"))
  # Omit all NA from the data frame
  HANES <- na.omit(HANES)
  # Observe the structure
  str(HANES)
  # Load the tidyverse library
  library(tidyverse)
```


```{r eval=TRUE, message=FALSE, echo = FALSE}
  # Load the package RCurl
  library(RCurl)
  # Import the HANES data set from GitHub; break the string into two for readability
  # (Please note this readability aspect very carefully)
  URL_text_1 <- "https://raw.githubusercontent.com/kannan-kasthuri/kannan-kasthuri.github.io"
  URL_text_2 <- "/master/Datasets/HANES/NYC_HANES_DIAB.csv"
  # Paste it to constitute a single URL 
  URL <- paste(URL_text_1,URL_text_2, sep="")
  HANES <- read.csv(text=getURL(URL))
  # Rename the GENDER factor for identification
  HANES$GENDER <- factor(HANES$GENDER, labels=c("M","F"))
  # Rename the AGEGROUP factor for identification
  HANES$AGEGROUP <- factor(HANES$AGEGROUP, labels=c("20-39","40-59","60+"))
  # Rename the HSQ_1 factor for identification
  HANES$HSQ_1 <- factor(HANES$HSQ_1, labels=c("Excellent","Very Good","Good", "Fair", "Poor"))
  # Rename the DX_DBTS as a factor
  HANES$DX_DBTS <- factor(HANES$DX_DBTS, labels=c("DIAB","DIAB NO_DX","NO DIAB"))
  # Omit all NA from the data frame
  HANES <- na.omit(HANES)
  # Observe the structure
  str(HANES)
  # Load the tidyverse library
  library(tidyverse)
```

Suppose we draw twenty people randomly and identify their gender:

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE}
  # Randomly sample 20 people and identify their gender
  random_gender <- sample(HANES$GENDER, 20, replace = FALSE)
  random_gender
```

```{r eval=TRUE, message=FALSE, warning=FALSE, echo=FALSE}
  # Randomly sample 20 people and identify their gender
  random_gender <- sample(HANES$GENDER, 20, replace = FALSE)
  random_gender
```

The first thing we can do is assume that we know the proportion of males and females in the population, and ask ourselves how likely it would have been to draw this sequence if we were right about the proportion. We will call the proportion of females $\gamma$ in this case.

Since there are only two categories, male and the female and each person in sampled independently the distribution being identical, we can calculate the probability of getting that sequence as the product of the probabilities of sampling each individual:

$$
L(\gamma \mid X) = \prod_{i=1}^{20}
\begin{cases}
    \gamma,      & \text{if } x_{i} = F \\
    1-\gamma,    & \text{if } x_{i} = M
\end{cases}
$$
Now we can use the above equation to calculate the probability of getting the specific sequence if the gender were equally distributed:

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE}
  # Assign probablities to the above seqence assuming men and women are equally distributed
  # in the general population
  probabilities = ifelse(random_gender == "F", 0.5, 1-0.5)
  # Likelihood is the product  of the probablities when samples are IID 
  likelihood = prod(probabilities)
  # Report the probability
  cat('Probability of observation when men and women are equally distributed:',likelihood)
```

```{r eval=TRUE, message=FALSE, warning=FALSE, echo=FALSE}
  # Assign probablities to the above seqence assuming men and women are equally distributed
  # in the general population
  probabilities = ifelse(random_gender == "F", 0.5, 1-0.5)
  # Likelihood is the product  of the probablities when samples are IID 
  likelihood = prod(probabilities)
  # Report the probability
  cat('Probability of observation when men and women are equally distributed:',likelihood)
```

If men are over-represented in the population, say, by 80%, then:

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE}
  # Assign probablities to the above seqence assuming men represent 80% of
  # the general population
  probabilities = ifelse(random_gender == "F", 0.2, 1-0.2)
  # Likelihood is the product  of the probablities when samples are IID 
  likelihood = prod(probabilities)
  # Report the probability
  cat('Probability of observation when men are represented 80%:',likelihood)
```

```{r eval=TRUE, message=FALSE, warning=FALSE, echo=FALSE}
  # Assign probablities to the above seqence assuming men represent 80% of
  # the general population
  probabilities = ifelse(random_gender == "F", 0.2, 1-0.2)
  # Likelihood is the product  of the probablities when samples are IID 
  likelihood = prod(probabilities)
  # Report the probability
  cat('Probability of observation when men are represented 80%:',likelihood)
```

Thus we can easily generalize this as a function of $\gamma$, and graph the likelihood as a function of any value of the proportion of females ranging from 0 to 1:

```{r eval=FALSE, message=FALSE, warning=FALSE, echo=TRUE}
  # Make a function that returns likelihoods based on the above sequence
  # and the gamma parameter
  likelihood = function(gamma, data = random_gender) {
    # Calculate the probabilities
    probabilities = ifelse(random_gender == "F", gamma, 1-gamma)
    # Calculate the product of probabilities, aka, likelihoods
    likelihood = prod(probabilities)
    # Return the likelihood
    return(likelihood)
  }
  # Make a vector of gammas
  gammas = seq(0.01, 0.99, by=0.01)
  # Compute likelihood through the function 
  likelihoods = sapply(gammas, likelihood)
  # Plot gammas vs. likelihoods
  ggplot(mapping = aes(x = gammas, y = likelihoods)) + geom_line(se=FALSE)
```

```{r eval=TRUE, message=FALSE, warning=FALSE, echo=FALSE}
  # Make a function that returns likelihoods based on the above sequence
  # and the gamma parameter
  likelihood = function(gamma, data = random_gender) {
    # Calculate the probabilities
    probabilities = ifelse(random_gender == "F", gamma, 1-gamma)
    # Calculate the product of probabilities, aka, likelihoods
    likelihood = prod(probabilities)
    # Return the likelihood
    return(likelihood)
  }
  # Make a vector of gammas
  gammas = seq(0.01, 0.99, by=0.01)
  # Compute likelihood through the function 
  likelihoods = sapply(gammas, likelihood)
  # Plot gammas vs. likelihoods
  ggplot(mapping = aes(x = gammas, y = likelihoods)) + geom_line(se=FALSE)
```

Likelihood is not a probability function. Also, given this function it can be used to identify the parameters (proportions or means) of the underlying population that is likely to maximize the chance of observing our data. This is the maximum likelihood estimate. Thus, maximum likelihood estimate is the value of the parameters of a given model that maximizes the likelihood function for the data. Mathematically,

$$
\gamma_{MLE} = \max_{\gamma}[L(\gamma \mid X)]
$$
In our example, it is the the ratio $\frac{\text{# of women}}{20}$, since it is the so many number of women we saw when we randomly sampled 20 people.

Determining maximum likelihood can become difficult as the complexity of the models increase. That is why we need optimization techniques and what are known as _cost functions_.

---

##### Cost functions

Cost functions are functions that maps a set of events into a number that represents the _cost_ of that event occurring. Mathematically it is defined as the negative log of likelihood.

$$
C(\gamma,X)=−\log[L(\gamma \mid X)]
$$
But why?

Lets watch this video for a good explanation:

---

<iframe width="560" height="315" src="https://www.youtube.com/embed/R4OlXb9aTvQ" frameborder="0" allowfullscreen></iframe>

---

There are other cost functions in areas such as motion planning and traveling salesman where different types of optimization techniques are used. For this tutorial we will use this cost function.

With this cost function as a negative log likelihood we can reformulate the cost of the given parameterization of the population and plot it:

---

**Theorem**: For a convex function defined on a convex set, any local minimum is also a global minimum.

_Proof_: Let $f:C \rightarrow \mathbb{R}$, be a convex function, on a convex set $C$. Suppose that $x$ is a local minimum of $f$ but not a global one. This means $\exists y \in C$ such that $f(y) < f(x)$. By definition of convexity for any $\alpha \in [0,1]$, we have,

$$ f(x+\alpha (y-x)) < f(x) + \alpha [f(y) - f(x)] \\
 < f(x) 
$$
since $f(y) < f(x)$. Hence, for any small ball $B(x;\epsilon)$ around $x$, there exists $z$ such that $f(z) < f(x)$. Therefore, $x$ is not a local minimum - a contradiction. QED.

---

#### References

1. An excellent tutorial here: [Introduction to Optimization](https://www.youtube.com/watch?v=PoTpmjvHxlg)