---
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<style>

body {
text-align: justify;}

figure figcaption {
    text-align: center;}
</style>


---

## Support Vector Machines  

---

#### Recorded Stream

---

<iframe width="560" height="315" src="https://www.youtube.com/embed/fTJB_rdNPJ4" frameborder="0" allow="autoplay; encrypted-media" allowfullscreen></iframe>

---

### Application to Gene Expression Data

---

We will use Kahn data set which consists of a number of tissue samples corresponding to four distinct types of small round blue cell tumors. We have both training data and testing data.

```{r eval=TRUE, message=FALSE}
  # Examine the dimension of the data
  library(ISLR)
  names(Khan)
  dim(Khan$xtrain)
  dim(Khan$xtest)
```

We will use SVM with the cost of 10. Note that we have more features than observations and so linear kernel should be sufficient.

```{r eval=TRUE, message=FALSE, warning=FALSE}
  # Make the Khan data frame
  dat=data.frame(x=Khan$xtrain, y=as.factor(Khan$ytrain))
  # Load svm library
  library(e1071)
  # Fit SVM model with cost parametr 10
  out=svm(y ~., data=dat, kernel="linear",cost=10)
  # Print the summary
  summary(out)
  # Make the confusion matrix
  table(out$fitted , dat$y)
```

We see there are no training errors. We now fit the model on the test data

```{r eval=TRUE, message=FALSE, warning=FALSE}
  # Make test data frame
  dat.te=data.frame(x=Khan$xtest, y=as.factor(Khan$ytest))
  # Make predcitions for the test data
  pred.te=predict(out, newdata=dat.te)
  # Make the confusion matrix table
  table(pred.te, dat.te$y)
```

We see the errors for the class labeled 3. There are 2 misclassifications.

---

#### Selected materials and references

An Introduction to Statistical Learning by Gareth James et al. Chapter 9: Lab exercises.

---