---
# title: "Lecture 03"
# author: "Kasthuri Kannan"
# date: "Sept 07, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

### Lecture 03 (Sept. 14, 2017)

---

#### R fundamentals 02: Advanced Data Types and Graphics

In this lecture we will consider other data types such as lists and data frames as well as graphics.

---

#### Factors

Factors are determined through categorical variables. What are categorical variables?

* Limited number of differing values.
* Belong to a certain category.
* For statistical analysis, R calls them factors.

##### Creating and manipulating factors

```{r eval=TRUE}
  # Create a blood group vector
  blood_group_vector <- c("AB", "O", "B+", "AB-", "O", "AB", "A", "A", "B", "AB-")
  # Create fatcors from the vector
  blood_group_factor <- factor(blood_group_vector)
  blood_group_factor
```

**Note** 

1. The levels are sorted alphabetically. 
2. No more quotation marks.

R encodes factors to integers for easier memory access and computations. This is done alphabetically. For example, A is assigned 1, AB is assigned 2 etc. This can be viewed by invoking the `str()` function:

```{r eval=TRUE}
  # Show the structure of the blood group factor
  str(blood_group_factor)
```

This can be over-ridden by specifying the levels argument for the `factor()` function.

```{r eval=TRUE}
  # Define another set of levels over-riding default
  blood_group_factor2 <- factor(blood_group_vector, levels = c("A", "B", "B+", "AB", "AB-", "O"))
  str(blood_group_factor2)
```

Renaming factors can be done using the `level()` function.

```{r eval=TRUE}
  # Define blood type
  blood_type <- c("B", "A", "AB", "A", "O")
  # Find the factors
  blood_type_factor <- factor(blood_type)
  blood_type_factor
  # Rename the factors
  levels(blood_type_factor) <- c("BT_A", "BT_AB", "BT_B", "BT_O")
  blood_type_factor
```

**Note** It is extremely **important** to follow the same order as the default order supplied by R. Otherwise, the result can be extremely confusing as the following exercise will show.

<br>
<span style="color:magenta">**Classwork**</span>: Rename the blood_type_factor in the above example as follows: 

`levels(blood_type_factor) <- c("BT_A", "BT_B", "BT_AB", "BT_O")` 

and justify the result of outputting `blood_type_factor`.
<br>

If you want to label the levels, it is always best to define the labels along with the levels like this -

```{r eval=FALSE}
  factor(blood_type, levels=c("A","B","AB","O"), labels=c("BT_A","BT_B","BT_AB","BT_O"))
```

##### Nominal vs. Ordinal factors

_Nominal factors_: Categorical variables that cannot be ordered, like blood group. For example, it doesn't make sense to say blood group A < blood group B.

_Ordinal factors_: Those categorical variables that can be ordered. For instance, tumor sizes. We can say T1 (tumor size 2cm or smaller) < T2 (tumor size larger than 2cm but smaller than 5 cm).

R provides us with the way to impose order on factors. Simply use the argument "ordered=TRUE" inside the factor function.

```{r eval=TRUE}
  # Specify the tumor size vectore
  tumor_size <- c("T1","T1","T2","T3","T1")
  # Set the order by specifying "ordered=TRUE"
  tumor_size_factor <- factor(tumor_size, ordered=TRUE, levels=c("T1","T2","T3"))
  # Print the results
  tumor_size_factor
  # Compare one factor vs the other
  tumor_size_factor[1] < tumor_size_factor[2]
```

<br>
<span style="color:magenta">**Classwork**</span>: Use the inequality operator (< or >) on a nominal category and print the output.
<br>

---

#### Lists

Recall vectors and matrices can hold only one data type, like integer or character. Lists can hold multiple R objects, without having to perform coercion.

```{r eval=TRUE}
  # Defining different data type as vector
  # Note coercion takes place
  vec <- c("Blood-sugar","High", 140, "Units", "mg/dL")
  vec
  # And as a list
  lst <- list("Blood sugar","High", 140, "mg/dL")
  lst
  # One can use the list function to see if something is a list
  is.list(lst)
```

Naming a list can be done through the `names()` function or specifying it in the list itself.

```{r eval=TRUE}
  # Define list
  lst <- list("Blood sugar","High", 140, "mg/dL")
  # Assign names and print
  names(lst) <- c("Fluid","Category","Value","Units")
  lst
  # Specify within the list
  list(Fluid="Blood sugar",Category="High", Value=140, Units="mg/dL")
  # For compact display use the str() function
  str(lst)
```  

**Note**: List can contain another list, or any number of nested lists.

##### Aceesing and extending lists

Difference between `[]` and `[[]]` is that, `[]` will return a list back and `[[]]` will return the elements in the list.

```{r eval=TRUE}
  # Define a list
  blood_test <- list(Fluid="Blood sugar",Category="High", Value=140, Units="mg/dL")
  # Make another list containing this list
  patient <- list(Name="Mike",Age=36,Btest=blood_test)
  # Access the first list
  patient[1]
  # Access the third "element" - which is actually a list itself.
  patient[[3]]
```

<br>
<span style="color:magenta">**Classwork**</span>: 

1. What does `patient[c(1,3)]` give us? Is it a list or elements?
2. What does `patient[[c(1,3)]` give us? Is it a list or elements? 
3. How about `patient[[c(3,1)]`? What is the difference? ( _Hint_: `patient[[c(1,3)]` is same as `patient[[1]][[3]]`).
<br>

Subsetting by names is super easy: just supply the name within brackets. For example, `patient["Name"]` or `patient[["Name"]]`. Subsetting by logicals will work only for returning the list. For instance, `patient[c(TRUE,FALSE)]`. It doesn't make sense to return the elements through logicals, i.e., `patient[[c(TRUE,FALSE)]]`.

Another cool way to access elements (just the same as using `[[]]`) is the use of `$` sign. However, to do this, the list should be named. For example, `patient$Name` will print the patient name.

`$` can also be used for extending lists:

```{r eval=FALSE}
  # Extend the list to include gender
  patient$Gender <- "Male"
  # This is same as using double brackets
  patient[["Gender"]] <- "Male"
  # Extend the blood test list to include the date of the test
  patient$Btest$Date <- "Sept.14"
```

<br>
<span style="color:magenta">**Classwork**</span>: How do you remove an element from a list?
<br>

---

#### Data frames

Datasets come with various shapes and sizes. Usually they constitute:

* Observations (eg. each row is an observation)
* Variables (eg. each column is a variable)
* Matrix consitute only one data type
* List is not practical 

Data frames can contain different types for each observation/row however each variable (or a column) should have a same data type. Usually data frames are imported - through CSV, or Excel etc. However, we can create a data frame as well.

```{r eval=TRUE}
  # Create name, age and logical vectors
  name <- c("Anne","James","Mike","Betty")
  age <- c(20,43,27,25)
  cancer <- c(TRUE,FALSE,FALSE,TRUE)
  # Form a data frame
  df <- data.frame(name,age,cancer)
  df
  # Create names function (like we did for vectors)
  names(df) <- c("Name","Age","Cancer_Status")
  df
  # Or specify inside data frame
  df <- data.frame(Name=name, Age=age, Cancer_Status=cancer)
  df
```

<br>
<span style="color:magenta">**Classwork**</span>: 

1. Examine the structure of the data frame.
2. What happens if one of the vectors have unequal length?
<br>

**Note**: Data frames store character vectors as factors. You can over-ride this as follows: 
`df <- data.frame(Name=name, Age=age, Cancer_Status=cancer, stringsAsFactors = FALSE)`

##### Manipulating data frames: Subsetting

```{r eval=TRUE}
  # Create name, age and logical vectors
  name <- c("Anne","James","Mike","Betty")
  age <- c(20,43,27,25)
  cancer <- c(TRUE,FALSE,FALSE,TRUE)
  # Form a data frame
  df <- data.frame(name,age,cancer)
  df
  # Subsetting by indices - works just like matrices
  df[1,2]
  # Subsetting by indices - one can use column names as well
  df[1,"age"]
  # Get the entire row/column - just like matrices
  # Get the second row
  df[2,]
  # Get the "cancer" column
  df[,"cancer"]
  # Get all 2nd and 3rd observation with "name"" and "cancer"" status
  df[c(2,3),c("name","cancer")]
```  

The only difference is when you specify a single number as index within `[]`. For matrices you get an element corresponding to the linear index but for data frame we'll get the column vector corresponding to the index. 

```{r eval=TRUE}
  # Print the second column
  df[2]
```

This is becuase data frames are made up of lists of vectors of equal length. Thus, single `[2]` will correspond to the second element in the list, which is a _vector_ of ages.

<br>
<span style="color:magenta">**Classwork**</span>: Test the operations of lists (like `[age] & [["age"]]`) to data frames.
<br>

##### Manipulating data frames: Extending

Extending a column is super easy. Since data frames are list of vectors one can just append a vector to the list. For instance, if we have a column of tumor size info like this for each patient: `c("T0","T3","T2","T0")`, the following code will append the vector.

```{r eval=TRUE}
  # Append tumor size vector
  df$tumor_size <- c("T0","T3","T2","T0")
  df
```

<br>
<span style="color:magenta">**Classwork**</span>: 

1. Use `cbind()` to append a vector of choice.
2. What happens if the length of the appending vector is greater than (or less than) row dimensions?
<br>

In contrast, extending a row (or observation) is not straight-forward. This is because observations may contain different data types. To add observations, make a new data frame and append:

```{r eval=TRUE}
  # Create a data frame
  tom <- data.frame(name="Tom", age=47,cancer="TRUE",tumor_size="T2")
  # And append
  df <- rbind(df,tom)
  df
```  

<br>
<span style="color:magenta">**Classwork**</span>:

1. Can you use the `list()` function instead of the data frame function in the above code?
2. What happens if you leave the arguments `name=`, `age=` etc. in the above code?
<br>

##### Manipulating data frames: Sorting

We can use the `order()` function to sort the entire data frame with respect to a particular column.

```{r eval=TRUE}
  # Rank the entries of a column, say "age"
  ranks <- order(df$age)
  # Sort the data frame according to the rank
  df[ranks,]
```

<br>
<span style="color:magenta">**Classwork**</span>:

1. What does `sort(df$age)` perform and how is it related to the ranks?
2. Sort the entries in descending order of the age. _(Hint: Question the function to find out more about the function in question)_.
<br>

---

#### Graphics

R has very strong graphical capabilities - primary reason why both industries and academics are interested.

* Create plots with code
* Replication and modification is easy
* Reproducibility!
* _graphics_ package loaded bu default produces great plots

#### Selected materials and references

[An Introduction to R](https://cran.r-project.org/doc/manuals/R-intro.pdf)