---
# title: "Lecture 03"
# author: "Kasthuri Kannan"
# date: "Sept 07, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

### R fundamentals 03: Elements of programming  (Sept. 19, 2017)

---

As the last part of R fundamentals series, we will study elements of R programming.

---

#### Conditional statements in R

R provides syntax to evaluate conditions. 

```{r eval=FALSE}
  if (condition) {
    expression
  }
```

As an example,

```{r eval=TRUE}
  # Assign x to -8
  x <- -8
  # Print if x is negative
  if (x < 0) {
    print("x is a negative number")
  }
```

<br>
<span style="color:magenta">**Classwork**</span>: Change x to 5 and re-run the above code. What does that print?

<br>

```{r eval=FALSE}
  if (condition) {
    expression 1
  }
  else {
    expression 2
  }
```


<br>
<span style="color:magenta">**Classwork**</span>: Evaluate the above code with the else statement printing "x is positive or zero", when it is.

<br>


```{r eval=FALSE}
  if (condition) {
    expression 1
  }
  else if {
    expression 2
  }
  else {
    expression 3
  }
```

R provides logical operators:

* AND with symbol &
* OR with symbol |
* NOT with symbol !

```{r eval=FALSE}
  # AND truth table
  TRUE  & TRUE   : TRUE
  TRUE  & FALSE  : FALSE
  FALSE & TRUE   : FALSE
  FALSE & FALSE  : FALSE
```

Example: `x <- 5` and `(x >3) & (x <8)` will evaluate to `TRUE`.

```{r eval=FALSE}
  # OR truth table
  TRUE  | TRUE   : TRUE
  TRUE  | FALSE  : TRUE
  FALSE | TRUE   : TRUE
  FALSE | FALSE  : FALSE
```

Example: `x <- 5` and `(x >5) | (x <8)` will evaluate to `FALSE`.

```{r eval=FALSE}
  # NOT truth table
  !TRUE  : FALSE
  !FALSE : TRUE
```

Example:  `x <- 5` and `!((x >5) | (x <8))` will evaluate to `TRUE`.

<br>
<span style="color:magenta">**Classwork**</span>: Create an R script that returns the max value of a vector x with length 3. Don’t use the aid of an auxiliary variable.
<br>

**Note**: R comes out of the loop as soon as the condition is met. It will **not** evaluate further conditions down the line even if they satisfy.

```{r eval=TRUE}
  # Assign x to 6
  x <- 6
  # Check if x is divisible by 2 or 3, if so print it is
  # Otherwise print it is not
  if (x %% 2 == 0) {
    print("x is divisioble by 2")
  } else if(x %% 3 == 0) {
    print("x is divisible by 3")
  } else {
    print("x is neither divisible by 2 nor 3")
  }
```

We can apply logical operators on vectors and matrices - works element wise:

`c(TRUE,TRUE,FALSE) & c(FALSE,TRUE,TRUE)` will evaluate to `FALSE TRUE FALSE`.

`c(TRUE,TRUE,FALSE) | c(FALSE,TRUE,TRUE)` will evaluate to `TRUE TRUE TRUE`.

Also, we have `&&` and `||` operators in R. The difference between single `&`and `&&`is that, `&&` will only look at the first elements of the vectors. Similarly for `||`. Thus,

`c(TRUE,TRUE,FALSE) && c(FALSE,TRUE,TRUE)` will evaluate to `FALSE` and 

`c(TRUE,TRUE,FALSE) | c(FALSE,TRUE,TRUE)` will evaluate to `TRUE`.

---

#### Selected materials and references

[An Introduction to R](https://cran.r-project.org/doc/manuals/R-intro.pdf)

---