---
# title: "Lecture 03"
# author: "Kasthuri Kannan"
# date: "Sept 07, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

### Lecture 03 (Sept. 14, 2017)

---

#### R Basics II

In this lecture we will consider other data structures, graphics and programming elements.

---

#### Factors

Factors are determined through categorical variables. What are categorical variables?

* Limited number of differing values.
* Belong to a certain category.
* For statistical analysis, R calls them factors.

##### Creating and manipulating factors

```{r eval=TRUE}
  # Create a blood group vector
  blood_group_vector <- c("AB", "O", "B+", "AB-", "O", "AB", "A", "A", "B", "AB-")
  # Create fatcors from the vector
  blood_group_factor <- factor(blood_group_vector)
  blood_group_factor
```

**Note** 

1. The levels are sorted alphabetically. 
2. No more quotation marks.

R encodes factors to integers for easier memory access and computations. This is done alphabetically. For example, A is assigned 1, AB is assigned 2 etc. This can be viewed by invoking the `str()` function:

```{r eval=TRUE}
  # Show the structure of the blood group factor
  str(blood_group_factor)
```

This can be over-ridden by specifying the levels argument for the `factor()` function.

```{r eval=TRUE}
  # Define another set of levels over-riding default
  blood_group_factor2 <- factor(blood_group_vector, levels = c("A", "B", "B+", "AB", "AB-", "O"))
  str(blood_group_factor2)
```

Renaming factors can be done using the `level()` function.

```{r eval=TRUE}
  # Define blood type
  blood_type <- c("B", "A", "AB", "A", "O")
  # Find the factors
  blood_type_factor <- factor(blood_type)
  # Rename the factors
  levels(blood_type_factor) <- c("BT_A", "BT_B", "BT_AB", "BT_O")
  blood_type_factor
```




#### Selected materials and references

[An Introduction to R](https://cran.r-project.org/doc/manuals/R-intro.pdf)