---
# title: "Lecture 03"
# author: "Kasthuri Kannan"
# date: "Sept 07, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

---

### R fundamentals 03: Elements of programming  (Sept. 19, 2017)

---

As the last part of R fundamentals series, we will study elements of R programming.

---

#### Conditional statements in R

R provides syntax to evaluate conditions. 

```{r eval=FALSE}
  if (condition) {
    expression
  }
```

As an example,

```{r eval=TRUE}
  # Assign x to -8
  x <- -8
  # Print if x is negative
  if (x < 0) {
    print("x is a negative number")
  }
```

<br>
<span style="color:magenta">**Classwork**</span>: Change x to 5 and re-run the above code. What does that print?

<br>

```{r eval=FALSE}
  if (condition) {
    expression 1
  }
  else {
    expression 2
  }
```


<br>
<span style="color:magenta">**Classwork**</span>: Evaluate the above code with the else statement printing "x is positive or zero", when it is.

<br>


```{r eval=FALSE}
  if (condition) {
    expression 1
  }
  else if {
    expression 2
  }
  else {
    expression 3
  }
```


<br>
<span style="color:magenta">**Classwork**</span>: Create an R script that returns the max value of a vector x with length 3. Don’t use the aid of an auxiliary variable.

<br>

**Note**: R comes out of the loop as soon as the condition is met. It will **not** evaluate further conditions down the line even if they satisfy.

```{r eval=TRUE}
  # Assign x to 6
  x <- 6
  # Check if x is divisible by 2 or 3, if so print it is
  # Otherwise print it is not
  if (x %% 2 == 0) {
    print("x is divisioble by 2")
  } else if(x %% 3 == 0) {
    print("x is divisible by 3")
  } else {
    print("x is neither divisible by 2 nor 3")
  }
```

---

#### Selected materials and references

[An Introduction to R](https://cran.r-project.org/doc/manuals/R-intro.pdf)

---